---
title: "Solution to Lab 1"
author: ''
date: "Due on 02/03 at 11:59 pm"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


**Instructions:** This lab report needs to be professional. Only report relevant and finalized code. Your writing should be concise and void of spelling errors. Use code chunk options to hide unnecessary messages/warnings. Your report should be reproducible. Reports that involve simulations need to have the random seed specified so that simulation results are reproducible. You are allowed to work on this lab assignment in groups of 2-3. You still need to submit an individual lab report if you do work in a group, and you need to list your collaborators. \vspace*{0.5cm}


**Question 1** In lecture it was demonstrated that baseball is a game of offense, pitching, and defense with a regression model that considered expected run differential as a function of explanatory variables OPS, WHIP, and FP. Do the following:

 - Fit a similar regression model with runs as the response variable. Report problems with this model. Investigate problematic residuals to discover what went wrong. Fix the problem with this model by adding categorical variable(s) to the list of explanatory variables. Briefly explain what went wrong. 

**Solution**

```{r message=FALSE}
library(Lahman)
library(tidyverse)
dat <- Teams %>%
    select(yearID, franchID, W, L, AB, H, X2B, X3B, HR, BB, HBP, SF,
                 HA, HRA, BBA, SOA, IPouts, FP, R, RA, G) %>%
    filter(yearID >= 1900) %>%
    replace_na(list(HBP = 0, SF = 0)) %>%
    mutate(RD = (R - RA) / (W + L), X1B = H - (X2B + X3B + HR)) %>%
    mutate(OBP = (H + BB + HBP)/(AB + BB + HBP + SF)) %>%
    mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR)/AB) %>%
    mutate(OPS = OBP + SLG) %>%
    mutate(WHIP = 3*(HA + BBA)/IPouts) %>%
    mutate(FIP = 3*(13*HRA + 3*BBA - 2*SOA)/IPouts)

mod_1a <- lm(R ~ OPS + WHIP + FP, data = dat)
summary(mod_1a)
## Q-Q plot seems weird
plot(mod_1a)
```

The plots appear to show normality except for several extreme negative residuals. The Q-Q plot shows the many observations drop at the left tail. 

```{r}
plot(mod_1a$residuals)
```

Several observations have large residuals

```{r}
unique(dat[which(abs(mod_1a$residuals) > 150),]$yearID)
```

Find the seasons that involve large residuals

```{r}
dat %>% filter(yearID %in% c(1918, 1981, 1994, 2020)) %>% select(G)
```

These seasons were all short seasons. 

```{r}
dat_game <- dat %>%
  mutate(game_level = ifelse(G <= 60, 1, 
                             ifelse(G <= 120, 2, 
                                    ifelse(G <= 140, 3, 
                                           ifelse(G <=160, 4, 5)))))
mod_1a2 <- lm(R ~ OPS + WHIP + FP + as.factor(game_level), data = dat_game)
summary(mod_1a2)
plot(mod_1a2)
```

As the model shows, the addition of a categorical variable for games played significantly improved the residuals in the model. 

WHIP and FP are defensive metrics and should have nothing to do with an offensive stat - Runs. WHIP has a negative coefficient, meaning that teams with weaker pitching will also score fewer runs, which checks out for the worst teams in the league. FP also has a negative coefficient, which means that teams who have a better defense will score fewer runs. This could illustrate the offense vs. defense aspect of constructing a lineup of position players.


 - We can significantly improve the regression model in the notes through a principled rescaling of OPS, WHIP, and FP. Split the Teams data frame by \texttt{yearID} and, for each year, create variables \texttt{OPSscale = OPS/avgOPS}, \texttt{WHIPscale = avgWHIP/WHIP}, and \texttt{FPscale = avgFP/FP} which require you to first create league average variables \texttt{avgOPS}, \texttt{avgWHIP}, and \texttt{avgFP}. Fit the linear regression model with runs differential as the response and explanatory variables \texttt{OPSscale}, \texttt{WHIPscale}, and \texttt{FPscale}, and report relevant output. Why does this model perform so much better than the model in the notes? Support your answer. Hint: functions \texttt{split}, \texttt{do.call}, and \texttt{lapply} are useful.


**Solution**

```{r}
avg_data <- dat %>%
group_by(yearID) %>% 
summarize(AB = sum(AB), H = sum(H), BB = sum(BB), HBP = sum(HBP), X2B = sum(X2B), 
          X3B = sum(X3B),HR = sum(HR), SF = sum(SF), HA = sum(HA), BBA = sum(BBA), 
          IPouts = sum(IPouts),avgFP = mean(FP), X1B = sum(X1B)) %>%
  mutate(OBP = (H + BB + HBP)/(AB + BB + HBP + SF)) %>%
    mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR)/AB) %>%
    mutate(avgOPS = OBP + SLG) %>%
    mutate(avgWHIP = 3*(HA + BBA)/IPouts) %>% ungroup() %>%
  select(yearID, avgWHIP, avgOPS, avgFP)

scale_data <- merge(dat, avg_data, by="yearID")
scale_data <- scale_data %>%
  mutate(WHIPscale = avgWHIP/WHIP) %>%
  mutate(OPSscale = OPS/avgOPS) %>%
  mutate(FPscale = avgFP/FP)

mod_1b <- lm(RD ~ OPSscale + WHIPscale + FPscale, data = scale_data)
summary(mod_1b)
plot(mod_1b)
```

This model performs better than the model in the notes because it provides context for the OPS, WHIP, and Fielding Percentage numbers based on the year and type of batted ball environment. By scaling each of these values by the league average, we can get a better understanding of how a team performed compared to the other teams that season.

**Question 2** Choose 3 batters and 3 pitchers that have played in at least 10 seasons and do the following: 

 - Display the seasonal statistics for these players. The following statistics should be included for batters (derivations of unconventional statistics are in parentheses): year, G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, SBpct (SB / (SB + CS)), BB, SO, OBP, SLG, OPS. The following statistics should be included for pitchers: year, W,  L, IPouts, H, ER, HR, BB, HBP, SO, ERA, WHIP, SOper9 (SO / IP * 9), SOperBB (SO / BB). These statistics can be found in or computed from statistics that are found in the \texttt{Batting} and \texttt{Pitching} dataframes in the \texttt{Lahman} package.

**Solution**

```{r}
batters <- Batting %>% 
  filter(playerID == "bondsba01" | playerID == "thomafr04" | playerID == "schmimi01") %>%
  mutate(X1B = H - (X2B + X3B + HR)) %>%
  mutate(SBpct = SB / (SB + CS)) %>%
  mutate(OBP = (H + BB + HBP)/(AB + BB + HBP + SF)) %>%
  mutate(SLG = (X1B + 2*X2B + 3*X3B + 4*HR)/AB) %>%
  mutate(OPS = OBP + SLG) %>%
  select(yearID, playerID, G, AB, R, H, X2B, X3B, HR, RBI, SB, CS, BB, SO, SBpct, OBP, SLG, OPS)

batters %>% filter(playerID == "bondsba01")
batters %>% filter(playerID == "thomafr04")
batters %>% filter(playerID == "schmimi01")

pitchers <- Pitching %>% 
  filter(playerID == 'maddugr01' | playerID == 'clemero02' | playerID == 'johnswa01') %>% 
  mutate(WHIP = (H + BB) / IPouts * 3) %>% 
  mutate(SOper9 = SO / IPouts / 3) %>%
  mutate(SOperBB = SO / BB) %>%
  select(yearID, playerID, W, L, IPouts, H, ER, HR, BB, HBP, SO, ERA, WHIP, SOper9, SOperBB)

pitchers %>% filter(playerID == 'maddugr01')
pitchers %>% filter(playerID == 'clemero02')
pitchers %>% filter(playerID == 'johnswa01')

```


 - Create career stat lines for each of the players that you selected. Be careful about how these statistics are calculated.

```{r}
career_batters <- Batting %>% 
  filter(playerID == "bondsba01" | playerID == "thomafr04" | playerID == "schmimi01") %>%
  group_by(playerID) %>%
  summarise(totalG = sum(G), totalAB = sum(AB), totalR = sum(R), totalH = sum(H), 
            totalX1B = sum(H - X2B - X3B - HR), 
            totalX2B = sum(X2B), totalX3B = sum(X3B), totalHR = sum(HR), 
            totalRBI = sum(RBI), totalSB = sum(SB), totalCS = sum(CS), 
            totalBB = sum(BB), totalSO = sum(SO), SBpct = totalSB / (totalSB + totalSB), 
            totalSF = sum(SF), totalHBP = sum(HBP), 
            totalOBP = (totalH + totalBB + totalHBP) / (totalAB + totalBB + totalHBP + totalSF), totalSLG = (totalX1B + 2*totalX2B + 3*totalX3B + 4*totalHR)/totalAB, 
            totalOPS = totalOBP + totalSLG)
career_batters

career_pitchers <- Pitching %>% 
  filter(playerID == 'maddugr01' | playerID == 'clemero02' | playerID == 'johnswa01') %>% 
  group_by(playerID) %>%
  summarise(totalG = sum(G), totalW = sum(W), totalL = sum(L), totalIPouts = sum(IPouts),
            totalER = sum(ER), totalHR = sum(HR), totalBB = sum(BB), totalHBP = sum(HBP),
            totalSO = sum(SO), totalERA = sum(ER)*27/sum(IPouts), 
            totalWHIP = (sum(H) + sum(BB)) *3/sum(IPouts), 
            totalSOper9 = 27 * sum(SO)/sum(IPouts), totalSOperBB = sum(SO)/sum(BB))
career_pitchers
```


 - Provide a plot for career trajectories for one batting and one pitching statistic of your choice. These are two separate graphics, one for the batters and one for the pitchers. The graphics that you produce should display the trajectories of the 3 batters and the 3 pitchers. Provide interesting commentary on your graphic.

```{r}
ggplot(batters, aes(yearID, H, colour = playerID)) + geom_line() + ggtitle("Hits per Season for Barry Bonds, Frank Thomas and Mike Schmidt ")
```

Mike Schmidt and Barry Bonds show their consistency in hits. 

```{r}
ggplot(pitchers, aes(yearID, SO, colour = playerID)) + geom_line() + ggtitle("Strikeouts per Season for Greg Maddux, Roger Clemens, and Walter Johnson")
```

Roger Clemens and Greg Maddux both are great pitchers, but clearly Roger Clemens is more powerful in strikeouts than Greg Maddux. 

**Question 3** Problem 2 on page 28 of Analyzing Baseball Data with R

**Solution**

```{r}
## a)

Bob_Gibson <- Pitching %>% filter(yearID==1968, playerID=='gibsobo01')

Bob_Gibson$CG / Bob_Gibson$GS

## b)
Bob_Gibson$SO / Bob_Gibson$BB

## c)
Bob_Gibson$IPouts / 3

## d)

(Bob_Gibson$H + Bob_Gibson$BB) / Bob_Gibson$IPouts *3

```


**Question 4** Problem 3 on page 29 of Analyzing Baseball Data with R 

```{r message=FALSE}
library(retrosheet)
data_4 <- getRetrosheet("game", 1964) %>% filter(Date == 19640621, VisTm == 'PHI', DblHdr == 1)

# a)
data_4$Duration

## Hour
floor(data_4$Duration / 60)

## minute
data_4$Duration %% 60
```



b) The attendance is 0 likely due to the doubleheader being played that day. 

```{r}
## c) 
data_4$VisD + data_4$VisT + data_4$VisHR
```
```{r}
## d)
OBP = (data_4$VisH + data_4$VisBB + data_4$VisHBP)/(data_4$VisAB + data_4$VisBB + data_4$VisHBP + data_4$VisSF)
OBP
```




